A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-03-29 CC/writeup.Rnw
added description about the layer. compiled after removing unneccesary print statements.

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, cache=FALSE, eval = TRUE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center',fig.show='hide')
@
<<readingdata, eval=TRUE, echo=FALSE>>=
setwd("~/Dropbox/ggplot2/") 
library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
names(blood$vertices)[1] <- 'label'
load("data/madmen.rda") # relative path should work for both of us
names(madmen$vertices)[1] <- 'label'
names(theme_elements$vertices)[1] <- 'label'
football$vertices$label <- as.factor(football$vertices$label)
football$vertices$value <- as.factor(football$vertices$value)
 email.net <- read.csv("data/email-data.csv")
 email.names <- data.frame(label = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to)))))

@@

\section{Technical Details}
\sctq{I have no idea exactly how detailed I should be. I'm going to attempt to be detailed without parsing each line in the code, but I do want to describe how everything works in detailed enough descriptions. I know you'll tell me if I'm too detailed or not detailed enough!}
\sct{
In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing.
  \subsection{Creating \texttt{geom\_net}}
  We constructed \texttt{geom\_net} using the drawing capabilities of the \texttt{geom\_point}, \texttt{geom\_line}, \emph{and \texttt{geom\_text}} functions that were already in \texttt{ggplot2}.  Using these capabilities, we were able to seamlessly incorporate our network geometry into the \texttt{ggplot2} structure.   We took our edges data frame and used that as our input to the \texttt{GeomSegment} draw function.  Then, on top of that, we plot the vertices by passing our vertex data frame to the \texttt{GeomPoint} draw function. \emph{Finally, we use the \texttt{GeomText} draw function to add the vertex and edge labels when the \texttt{vlabel} and \texttt{elabel} arguments, respectively, are true. *insert more detail here after figuring it out*}
} \sctq{This doesn't seem very long, but I already feel like there's too much detail! Should I add more?}

>>>>

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, cache=FALSE, eval = TRUE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center',fig.show='hide')
@
<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
names(blood$vertices)[1] <- 'label'
load("data/madmen.rda") # relative path should work for both of us
names(madmen$vertices)[1] <- 'label'
names(theme_elements$vertices)[1] <- 'label'
football$vertices$label <- as.factor(football$vertices$label)
football$vertices$value <- as.factor(football$vertices$value)
 email.net <- read.csv("data/email-data.csv")
 email.names <- data.frame(label = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to)))))

@@

\section{Technical Details}
\sctq{I have no idea exactly how detailed I should be. I'm going to attempt to be detailed without parsing each line in the code, but I do want to describe how everything works in detailed enough descriptions. I know you'll tell me if I'm too detailed or not detailed enough!}
\sct{
In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing.
  \subsection{Creating \texttt{geom\_net}}
  We constructed \texttt{geom\_net} using the drawing capabilities of the \texttt{geom\_point}, \texttt{geom\_line}, \emph{and \texttt{geom\_text}} functions that were already in \texttt{ggplot2}.  Using these capabilities, we were able to seamlessly incorporate our network geometry into the \texttt{ggplot2} structure.   We took our edges data frame and used that as our input to the \texttt{GeomSegment} draw function.  Then, on top of that, we plot the vertices by passing our vertex data frame to the \texttt{GeomPoint} draw function. \emph{Finally, we use the \texttt{GeomText} draw function to add the vertex labels when the \texttt{vlabel} argument is TRUE. We added an if/else statement in the \texttt{draw} function which created a labeling layer which is \texttt{NULL} if \texttt{vlabel = FALSE}, which is the default. We then created a new data frame for the label layer. This allows the vertex properties to be propogated through to the labels, including size and color.}
} \sctq{This doesn't seem very long, but I already feel like there's too much detail! Should I add more?}
